--
title: "Introduction to R for Biostatistics 3"
output: html_document
---

```{r global_options, include=FALSE}
library(dplyr)
library(ggplot2)
```

#In-class worksheet 3

**X of March, 2016**

In this worksheet you will learn how to graphically display your data. R has a very wide range of different plotting routines, which allow you to generate publication style figures.

## Plotting

###Instructions

histograms + grouped histograms

meaning full axis labels

```{r}
salaries <- read.csv("D:/lab/teaching/statistics2016//R_for_biostatistics/salary.csv")

## Basic histogram from the vector "rating". Each bin is .5 wide.
## These both result in the same output:
ggplot(salaries, aes(x=sl)) + geom_histogram(bins=10)
# qplot(salaries$sl, binwidth=.5)

```


We can also overlay with a kernel density estimate (ie. smoothed)


```{r}
# Histogram overlaid with kernel density curve
ggplot(salaries, aes(x=sl)) + 
    geom_histogram(aes(y=..density..),      # Histogram with density instead of count on y-axis
                   bins=10,
                   colour="black", fill="white") +
    geom_density(alpha=.2, fill="#FF6666",adjust=.5)  # Overlay with transparent density plot


```
adding mean

```{r}
ggplot(salaries, aes(x=sl)) +
    geom_histogram(bins=10, colour="black", fill="white") +
    geom_vline(aes(xintercept=mean(sl, na.rm=T)),   # Ignore NA values for mean
               color="red", linetype="dashed", size=1)

```

grouped histograms

```{r}
# Overlaid histograms
ggplot(salaries, aes(x=sl, fill=sx)) +
    geom_histogram(bins=10, alpha=.5, position="dodge")
```

faceted histograms

```{r}
# Overlaid histograms
ggplot(salaries, aes(x=sl)) +
    geom_histogram(bins=10, alpha=.5) + 
    facet_grid(rk ~ . )
```




scatterplots + colored scatter plots







boxplots
```{r}
# A basic box with the conditions colored
ggplot(salaries, aes(x=rk, y=sl, fill=rk)) + geom_boxplot() + guides(fill=FALSE)

```




##Assignments

