--
title: "Introduction to R for Biostatistics 4"
output: html_document
---

```{r global_options, include=FALSE}
library(plyr)
library(dplyr)
library(ggplot2)
library(Rmisc)

set.seed(1)    # this makes everything reproducible

```

#In-class worksheet 4

**29 Feb to 4 Mar, 2016**

In this worksheet you will learn to generate data from different distributions and how to graphically assess whether a dataset is normally distrbuted. Please answer the questions in the document.

## Generating data from distributions

### Instructions

For many applications, it is important to be able to generate random numbers from specified distributions. R contains functions for handling many distributions, also of course for the normal distribution. Indeed, there is a whole family of functions for generating data from a normal distribution (*rnorm*), to obtain the normal density function (*dnorm*) and the cumulative density function (*pnorm*). Let's start with generating some data: 


```{r}

m = 5   # mean
s = 2   # sd

x <- rnorm(50, mean = m, sd = s)      # generate 50 samples form a normal distribution with mean 5 and sd 2
data = data.frame(x)

ggplot(data, aes(x=x)) +     
  stat_bin(aes(y=..count../sum(..count..))) +       # histogram of the samples with 20 bins
  geom_vline(aes(xintercept=mean(x, na.rm=T)),        # compute mean of the samples and add to plot as vertical line
             color="green", linetype="dashed", size=.75) + 
  geom_vline(aes(xintercept=m),                 # add true mean to plot as vertical line
             color="red", linetype="dashed", size=.75) + 
  stat_function(fun = dnorm, colour = "red", 
                arg = list(mean = m, sd=s)) +   # add the gaussian density with the true parameters
  
  xlab("x") + ylab("Fraction") +   
  theme_grey(base_size=14)  + 
  xlim(m-3*s, m+3*s)
```

Is the sample mean the same as the true mean? 

> Provide your answer here.

Run the cell multiple times. Does the histogram always look the same and lead to the same mean?

> Provide your answer here.

Play with the parameters of the distribution to observe what changes (mind the scale of the x-axis!).

> Provide your answer here.

Let's look at the relationship between the sample size and the variability of the mean in a bit more detail. We will now generate 1000 samples of the same mean and sd for each of a set of sample sizes. For each of these, we will not look at the distribution of the data points, but at the distribution of the *means* of each sample. 

```{r}

m = 5   # mean
s = 2   # sd
n = c(10, 20, 50, 200, 500)   # we will compute the mean from samples of these sizes 
rep = 1000           # for each sample size, we will do that 1000 times


# this loop generates the data - try to understand it
x = numeric(length(n)*rep)
nsamp = numeric(length(n)*rep)

for(i in 1:length(n)){
  for(j in 1:rep){
    x[(i-1)*rep+j] = mean(rnorm(n[i],mean=m,sd=s))
    nsamp[(i-1)*rep+j] = n[i]
  }
}

data = data.frame(x,nsamp)


# we create faceted histograms of the data
ggplot(data, aes(x=x)) +
  geom_histogram(aes(y=..density..),bins=50,alpha=.5, position="stack") + 
  facet_grid(nsamp ~ . ) +                       
  xlab("Sample mean") + ylab("Density") + 
  theme_grey(base_size=14)  

```

Are there any samples whos mean is larger than six for samples of size 10? What about samples of size 500?

> Provide your answer here.

Describe how the with of the distribution of means changes with the sample size.  

> Provide your answer here.

The width of this distribution is called the standard error of the mean. This quantity is the standard deviation of the distribution of means and is defined as $SEM=\frac{\sigma}{\sqrt{N}}$. We can check graphically that this formula is correct:

```{r}

# compute summary statistics about the distribution of means
data_sum <- summarySE(data = data, measurevar = "x", groupvars = c("nsamp"))    

semfun <- function(n,s) {s/sqrt(n)}   # this is the SEM formula

ggplot(data_sum, aes(x=nsamp, y=sd)) +  # the sd of the distribution of means is the SEM
  geom_point() +      
  stat_function(fun=semfun, args = list(s=2) ) +  # here we evaluate the SEM formula     
  xlab("Number of samples") + ylab("SEM") + 
  theme_grey(base_size=14)  
```

This looks reassuring. The points are all on the line. What does this mean?

> Provide your answer here.

We are now able to generate samples from a Normal distribution. In addition, we are interested in the question, what fraction of samples lies below a certain value. We can easily obtain this number from a *cumulative histogram*: 

```{r}

m = 5   # mean
s = 2   # sd

x <- rnorm(50, mean = m, sd = s)      # generate 50 samples form a normal distribution with mean 5 and sd 2
data = data.frame(x)

ggplot(data,aes(x=x)) +
  geom_histogram(aes(y=cumsum(..count..)), bins = 20) + 
  ylab("Cumulative count")

```

In the sample, roughly how many samples are smaller than 2.5?

> Provide your answer here.

How can you use this plot to graphically estimate the mean? Add the required lines to the plot (use the *+geom_hline* command).

> Provide your answer here.

Such a function is also defined for the Normal distribution, not only for samples from it. There it is called the *cumulative density function* or *distribution function*. It is very useful, because it let's you assess the percentiles of certain values quite easily. This is how it looks (for comparison, the density is shown with dashes):

```{r}

m = 5   # mean
s = 2   # sd

x <- -5:-1:15      
data = data.frame(x)

ggplot(data,aes(x=x)) +
  stat_function(fun = pnorm, colour = "red", 
                arg = list(mean = m, sd=s)) +   # add the gaussian density with the true parameters
  stat_function(fun = dnorm, colour = "red", 
                arg = list(mean = m, sd=s), linetype="dashed") +   # add the gaussian density with the true parameters
  ylab("Cumulative density")

```

If you are given a cumulative density function, you can find out the mean and the standard deviation of a dataset. How? 

> Provide your answer here.

For many applications, it is important to diagnose whether a given dataset is from a Normal distribution or not. One graphical tool to do this are *qq-plots*.

```{r}

m = 5   # mean
s = 2   # sd

x <- rnorm(50, mean = m, sd = s)      # generate 50 samples form a normal distribution with mean 5 and sd 2
data = data.frame(x)


ggplot(data,aes(sample=x)) +
  stat_qq()

ggplot(data, aes(sample = data$column1)) + stat_qq(color="firebrick2", alpha=1) + geom_abline(intercept = mean(data$column1), slope = sd(data$column1))

```


###Assignments



## Assessment of normality



