---
title: "Introduction to R for Biostatistics - Day 1"
output: html_document
---

##In-class worksheet 1

**X of March, 2016**

We will use R in this class mainly via R Markdown documents. R Markdown documents are documents that combine text, R code, and R output, including figures. They are a great way to produce self-contained and documented statistical analyses.

In this first worksheet, you will learn how to do some basic markdown editing and some simple data manipulation in R. After you have made a change to the document, press "Knit HTML" in R Studio and see what kind of a result you get.

Edit only below this line.

---------------------------------------------
## 1. Basic Markdown editing
Try out basic R Markdown features. You may know the Markdown language from editing wikis. Write some text that is bold, and some that is in italics. Make a numbered list and a bulleted list. Make a nested list. 

You can get bold text by enclo

This text is **bold.**

This text is *in italics.*

A numbered list:

1. Item 1
2. Item 2
3. Item 3

A bulleted list:

- Item 1
- Item 2
- Item 3

A nested list:

1. Item 1
    - Item 1.1. Note that 4 spaces are required for the nesting to work properly.
    - Item 1.2
2. Item 2



Block quote:

> "If we knew what it was we were doing, it would not be called research, would it?" --- Albert Einstein


Wir werden lernen, mit R grundlegende biostatistische Analysen durchzuführen. Dafür benötigen wir heute ersteinmal einige Grundlagen.

```{r}
summary(cars)
```

You can also embed plots, for example:

```{r}
plot(cars)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
